\vspace{1.0cm} \LARGE \bf Revealing the competitive dynamics of influenza viruses through evolutionary cartography \\ or \\ Charting antigenic drift and evolutionary success of seasonal influenza lineages

### 2013-01-29 flux.tex
Include revisions based on Philippe's suggestions.

Previously, the antigenic and genetic patterns of influenza evolution have been analyzed essentially in isolation. 
An antigenic map is constructed from a panel of HI measurements, and a phylogenetic tree is constructed from sequence data. 
However, the opportunity for a combined approach exists as both the antigenic map and the phylogenetic tree often contain many of the same isolates. 
Here, we implement a flexible Bayesian approach to characterize jointly the antigenic and genetic evolution of the influenza virus population. 
We apply this approach to investigate the dynamics of A/H3N2, A/H1N1, B/Vic and B/Yam viruses.

@@

We suggest that the larger signal of positive selection for antigenic diffusion in influenza A relative to influenza B could be due to either a greater fundamental reproductive number $R_0$ or to pleiotropic effects associated with antigenic change.
In the latter case, we expect that many antigenic variants may be at an inherent disadvantage against other strains from decreased protein function \cite{Kaverin04,Rudneva12} or other effects, even though they benefit from a transmission advantage mediated by host immunity.
These mutations may not be promoted by natural selection to the same extent as antigenic variants that maintain functional equivalence.
Still, it is clear that more work is necessary to fully clarify the the mechanistic underpinnings of the relative evolutionary success of different influenza lineages.

@@

The statistical framework presented here represents a baseline to which further advancements in modeling antigenic phenotype and evolution may be made.
For example, our likelihood-based model facilitates the inclusion of possible covariates affecting immunological titer, which could include experimental factors such as red blood cell type used in the HI assay \cite{Lin12} and whether oseltamivir is included in the HI reaction \cite{Lin10}.
Additionally, this framework should be ideally suited to uncovering genetic determinants of antigenic change, as both the sequence state and antigenic location of internal nodes in the phylogeny may be estimated.
In this fashion, it should be possible to correlate sequence substitutions directly to antigenic diffusion.

@@

\subsection*{Likelihood functions}

@@

We calculate the overall likelihood by multiplying probabilities of individual measurements
\begin{equation} 
	L(\viruses,\sera) = \prod_{(i,j) \in \cal I} f(H_{ij}),
\end{equation}
using probability functions $\point$, $\threshold$ and $\interval$ as appropriate.
We begin by assuming independent, diffuse normal prior on virus and serum locations
\begin{eqnarray}
	\virus_i &\sim& \mathrm{Normal}(\boldsymbol{\mu},\boldsymbol{\Sigma}) \nonumber \\
	\serum_j &\sim& \mathrm{Normal}(\boldsymbol{\mu},\boldsymbol{\Sigma}),
\end{eqnarray}
with $\boldsymbol{\mu} = (0,\ldots,0)^{\prime}$ and $\boldsymbol{\Sigma}$ is a diagonal matrix with diagonal elements all equal to $10000$.

@@

\subsection*{Location prior}

As presented, virus and serum locations $\viruses$ and $\sera$ give a likelihood of observing the HI data $\mathbf{H}$.
However, distances between viruses and sera represent only a local description of antigenic space and do not provide a global picture.
An example of this phenomenon is shown in Figure~\ref{schematic_map}.
In this case it is impossible to determine from the HI data at hand whether the blue and yellow viruses are antigenically similar or antigenically divergent. 
This presents an issue of model identifiability; multiple antigenic maps give the same likelihood of observing the serological data.
In order to achieve more interpretable antigenic locations we impose a weak prior on global locations.
In influenza, it's clear that antigenic distance between strains increases with time \cite{Smith04,Cai10}.
To capture this, we replace our previous diffuse prior with an informed prior in which the expected location of viruses and sera increases with date of sampling along dimension one, and each virus and serum location follows an independent normal distribution centered around this temporal expectation, so that
\begin{eqnarray}
	x_{i1} &\sim& \beta \, t_i + \normal(0, \virussd^2) \nonumber \\
	y_{j1} &\sim& \beta \, t_j + \normal(0, \serumsd^2),
\end{eqnarray}
where $t$ is the difference between the date of the indexed virus or serum and the date of the earliest sampled virus or serum, and other dimensions follow $x_{im} \sim \normal(0, \virussd^2)$ and $y_{jm} \sim \normal(0, \serumsd^2)$ for $m\ge2$.
Thus, this model assumes that virus and serum locations drift in a line across the antigenic map at rate $\beta$.
The parameter $\virussd$ determines the breadth of the cloud of virus locations at each point in time, while $\serumsd$ determines the breadth of the cloud of serum locations.

%%% schematic_map %%%
\begin{figure}[tb]
	\centering		
	\includegraphics[width=0.95\textwidth]{figures/schematic_map}
	\caption{\textbf{Schematic antigenic map with three viruses and two sera.} 
	Virus 1 is shown in blue, virus 2 is shown in red and virus 3 is shown in yellow.
	Virus isolates are represented by filled circles, sera raised against viruses are shown as open circles and map distances $\delta_{ij}$ are shown as solid lines connecting viruses and sera.
	Sera from virus 1 is compared against viruses 1 and 2, while sera from virus 2 is compared against viruses 2 and 3.
	Left and right panels represent cartographic models that would give equal likelihoods to a set of serological data $\{H_{11},H_{21},H_{22},H_{32}\}$.
	} 
	\label{schematic_map} 
\end{figure}

\subsection*{Diffusion model of antigenic evolution}

@@

Phylogenetic trees were sampled for A/H3N2, A/H1N1, B/Vic and B/Yam using BEAST \cite{BEAST17} and incorporated the SRD06 nucleotide substitution model \cite{Shapiro06}, a coalescent demographic model with constant effective population size and a strict molecular clock across branches.
MCMC was run for 60 million steps and trees were sampled every 50,000 steps after allowing a burn-in of 10 million steps, yielding a total sample of 2000 trees.
These trees were treated as a discrete set of possibilities when subsequently sampled in the BMDS analysis \cite{Pagel04}.

@@

There is some difficulty summarizing posterior cartographic samples, as sampled virus and serum locations represent only relative quantities, and because of this, over the course of the MCMC, virus locations may shift.
Our prior on virus and serum locations removes much of this issue, orienting the antigenic map along dimension 1 and fixing it to begin at the origin.
However, local isometries are often still a problem.
For example, in A/H3N2 the Beijing/89 cluster may shift back and forth from being above other viruses on dimension 2 to being below other viruses on dimension 2.
These isometries may be local, i.e.\ Beijing/89 and Beijing/92 may flip while leaving Texas/77 in place.
Consequently, it is difficult to fully align MCMC samples using Procrustes analysis.
For the present study, we take a simple approach and sample a single MCMC step and visualize the antigenic locations at this state (Figures~\ref{map}, Figure~\ref{drift}).
Then, for specific quantities of interest, like rate of antigenic drift and rate of diffusion at different points along the phylogeny, we calculate this quantity across MCMC samples to yield an expectation and a credible interval.
This approach accurately characterizes uncertainty that may be hidden in an analysis of a single antigenic map.

>>>>

Previously, the antigenic and genetic patterns of influenza evolution have been analyzed essentially in isolation. 
An antigenic map is constructed from a panel of HI measurements, and a phylogenetic tree is constructed from sequence data. 
However, the opportunity for a combined approach exists as both the antigenic map and the phylogenetic tree often contain many of the same isolates. 
Here, we implement a flexible Bayesian approach to characterize jointly the antigenic and genetic evolution of the influenza virus population. 
We apply this approach to investigate the dynamics of A/H3N2, A/H1N1, B/Vic and B/Yam viruses, and, for the first time, present detailed reconstructions of the antigenic dynamics of all four circulating influenza clades.

@@

We suggest that the larger signal of positive selection for antigenic diffusion in influenza A relative to influenza B could be due to either a greater fundamental reproductive number $R_0$ or to pleiotropic effects associated with antigenic change.
In the latter case, we expect that many antigenic variants may be at an inherent disadvantage against other strains from decreased protein function \cite{Kaverin04,Rudneva12} or other effects, even though they benefit from a transmission advantage mediated by host immunity.
These mutations may not be promoted by natural selection to the same extent as antigenic variants that maintain functional equivalence.
Additionally, other work has shown strong pleiotropic constraints in the evolution of oseltamivir resistance in A/H1N1 \cite{Bloom10}.
Still, it is clear that more work is necessary to fully clarify the the mechanistic underpinnings of the relative evolutionary success of different influenza lineages.

@@

The statistical framework presented here represents a baseline to which further advancements in modeling antigenic phenotype and evolution may be made.
For example, our likelihood-based model facilitates the inclusion of possible covariates affecting immunological titer, which could include experimental factors such as red blood cell type used in the HI assay \cite{Lin12} and whether oseltamivir is included in the HI reaction \cite{Lin10}.
Additionally, this framework should be ideally suited to uncovering genetic determinants of antigenic change, as both the sequence state and antigenic location of internal nodes in the phylogeny may be estimated.
In this fashion, it should be possible to correlate sequence substitutions directly to antigenic diffusion.
Here, methods to efficiently calculate branch- and site-specific codon substitutions may be necessary to easily correlate sequence substitutions with antigenic change \cite{Lemey12}.

@@

@@

We calculate the overall likelihood by multiplying probabilities of individual measurements
\begin{equation} 
	L(\viruses,\sera) = \prod_{(i,j) \in \cal I} f(H_{ij}),
\end{equation}
using probability functions $\point$, $\threshold$ and $\interval$ as appropriate.
We begin by assuming independent, diffuse normal priors on virus and serum locations
\begin{eqnarray}
	\virus_i &\sim& \mathrm{Normal}(\boldsymbol{\mu},\boldsymbol{\Sigma}) \nonumber \\
	\serum_j &\sim& \mathrm{Normal}(\boldsymbol{\mu},\boldsymbol{\Sigma}),
\end{eqnarray}
where $\boldsymbol{\mu} = (0,\ldots,0)^{\prime}$ and $\boldsymbol{\Sigma}$ is a diagonal matrix with diagonal elements all equal to $10000$.

@@

\subsection*{Drift model of antigenic evolution}

As presented, multiple configurations of virus and serum locations $\viruses$ and $\sera$ will give the same likelihood of an observed data matrix $\mathbf{H}$.
An example of this phenomenon is shown in Figure~\ref{schematic_map}.
In this case, it is impossible to determine from the HI data at hand whether the blue and yellow viruses are antigenically similar (Figure~\ref{schematic_map}A) or antigenically divergent (Figure~\ref{schematic_map}B). 
This presents an issue of model identifiability, where absolute, as opposed to relative, antigenic locations cannot be determined from the observing the serological data alone.
Thus, in order to achieve more a interpretable model we impose a weak prior on global locations.
In influenza, it's clear that antigenic distance between strains increases with time \cite{Smith04,Cai10}.
To capture this, we replace our previous diffuse prior with an informed prior in which the expected location of viruses and sera increases with date of sampling along dimension one, and each virus and serum location follows an independent normal distribution centered around this temporal expectation, so that
\begin{eqnarray}
	x_{i1} &\sim& \beta \, t_i + \normal(0, \virussd^2) \nonumber \\
	y_{j1} &\sim& \beta \, t_j + \normal(0, \serumsd^2),
\end{eqnarray}
where $t$ is the difference between the date of the indexed virus or serum and the date of the earliest sampled virus or serum, and other dimensions follow $x_{im} \sim \normal(0, \virussd^2)$ and $y_{jm} \sim \normal(0, \serumsd^2)$ for $m\ge2$.
Thus, this model assumes that virus and serum locations drift in a line across the antigenic map at rate $\beta$.
The parameter $\virussd$ determines the breadth of the cloud of virus locations at each point in time, while $\serumsd$ determines the breadth of the cloud of serum locations.

%%% schematic_map %%%
\begin{figure}[tb]
	\centering		
	\includegraphics[width=0.95\textwidth]{figures/schematic_map}
	\caption{\textbf{Schematic antigenic map with three viruses and two sera.}
	(A) Map with virus 1 and virus 3 antigenically similar.
	(B) Map with virus 1 and virus 3 antigenically divergent.
	Virus 1 is shown in blue, virus 2 is shown in red and virus 3 is shown in yellow.
	Virus isolates are represented by filled circles, sera raised against viruses are shown as open circles and map distances $\delta_{ij}$ are shown as solid lines connecting viruses and sera.
	Sera from virus 1 is compared against viruses 1 and 2, while sera from virus 2 is compared against viruses 2 and 3.
	Configurations (A) and (B) represent cartographic models that would give equal likelihoods to a set of serological data $\{H_{11},H_{21},H_{22},H_{32}\}$.
	} 
	\label{schematic_map} 
\end{figure}

\subsection*{Phylogenetic diffusion model of antigenic evolution}

@@

Phylogenetic trees were estimated for A/H3N2, A/H1N1, B/Vic and B/Yam using BEAST \cite{BEAST17} and incorporated the SRD06 nucleotide substitution model \cite{Shapiro06}, a coalescent demographic model with constant effective population size and a strict molecular clock across branches.
MCMC was run for 60 million steps and trees were sampled every 50,000 steps after allowing a burn-in of 10 million steps, yielding a total sample of 2000 trees.
These trees were treated as a discrete set of possibilities when subsequently sampled in the BMDS analysis \cite{Pagel04}.
However, it would be possible to jointly sample from sequence data and serological data using these methods.

@@

There is some difficulty summarizing posterior cartographic samples, as sampled virus and serum locations represent only relative quantities, and because of this, over the course of the MCMC, virus locations may shift.
Our prior on virus and serum locations removes much of this issue, orienting the antigenic map along dimension 1 and fixing it to begin at the origin.
However, local isometries are often still a problem.
For example, in A/H3N2 the Beijing/89 cluster may shift back and forth from being above other viruses on dimension 2 to being below other viruses on dimension 2.
These isometries may be local, i.e.\ Beijing/89 and Beijing/92 may flip while leaving Victoria/75 in place.
Consequently, it is difficult to fully align MCMC samples using Procrustes analysis.
For the present study, we take a simple approach and sample a single MCMC step and visualize the antigenic locations at this state (Figures~\ref{map}, Figure~\ref{drift}).
Then, for specific quantities of interest, like rate of antigenic drift and rate of diffusion at different points along the phylogeny, we calculate this quantity across MCMC samples to yield an expectation and a credible interval.
This approach accurately characterizes uncertainty that may be hidden in an analysis of a single antigenic map.
